Learning pleiotropic signatures to construct a causal map of the human phenome

### 2017-07-29 docs/manuscript.tex
updated manuscript

Gibran Hemani\(^1\), Jack Bowden\(^1\), Philip Haycock\(^1\), Jie
Zheng\(^1\), Oliver Davis\(^1\), Peter Flach, Tom Gaunt\(^{1,*}\),
George Davey Smith\(^{1,*}\)

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
  School of Social and Community Medicine, Bristol, UK
\end{enumerate}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this. Here we show that the
mechanisms that underlie horizontal pleiotropy are numerous, and that
instrument selection will be increasingly liable to selecting invalid
instruments as GWAS sample sizes continue to grow. We have developed a
mixture of experts machine learning approach (MR-MoE 1.0) that improves
on both power and false discovery rates over unselective use of any
existing methods. Using the approach, we systematically estimated the
causal effects amongst 2407 phenotypes, generating a working draft of
the causal map of the human phenome. Almost 90\% of causal estimates
indicated some level of horizontal pleiotropy. All results are organised
into a publicly available graph database (\url{http://eve.mrbase.org}).

@@

The use of summary data in two-sample MR is described in detail
elsewhere (11). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of MR (that the instrument associates with the exposure) is
generally satisfied. We collect their effect sizes, standard errors and
effect alleles for the association with the exposure trait. These can be
obtained manually from complete summary data or from curated lists of
significant GWAS associations. Next, the effects of those SNPs on the
outcome need to be obtained, typically necessitating complete summary
data because these SNPs will generally have not have reached genome-wide
significance for other traits (and therefore be present in curated
catalogues). At its most simple implementation, the regression of the
SNP-exposure effect sizes against the SNP-outcome effect sizes, with
greater weight afforded to those SNPs with smaller SNP-outcome standard
errors, provides the estimate of the causal effect of the exposure on
the outcome. This is known as the inverse-variance weighted (IVW)
method.

@@

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(8,12,25), based on the notion that outlying instruments are likely to
be exhibiting horizontal pleiotropy. Conversely, we primarily developed
Steiger filtering to identify those instruments that are likely to be
arising due to reverse cause.

The Steiger test is applied to each variant in turn and we exclude any
\(g_{A}\) for which \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\), indicating
that it is unlikely to primarily associate with \(B\) relative to \(A\).
Similarly, for SNPs that influence confounders of \(A\) and \(B\) or
exhibit horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant. Hence we also exclude any \(g_{A}\) for which the Steiger
Z-test has \(p > 0.05\).

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
\(p\) is the allele frequency of the SNP in the population (26). \(p\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\). We
exclude any SNPs that do not have the correct causal direction for the
hypothesis, and do not have Steiger test \(p < 0.05\).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). In the context of this
analysis, each method is considered to be an `expert', taking a set of
SNP-exposure and SNP-outcome effect sizes and their standard errors as
inputs. Our objective is to select the expert most likely to be correct
for a specific MR analysis. A overview of the approach is shown in
Figure 2.

The mixture of experts method is a machine learning approach (27) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that dataset.
The metrics are a mixture of regression diagnostics and are described in
Supplementary table 3.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated datasets. We trained the gating function using random forest
learning algorithms that seek to identify sectors of the parameter space
that are most likely to return accurate estimates for a particular
expert. Figure 2a illustrates how the gating function is trained using
simulated data. New datasets can then be applied to the trained mixture
of experts (MoE) model (Figure 2b). We call this implementation MR-MoE
1.0.

It should be noted that in the original hierarchical mixture of experts
approach of (27) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (28--30) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the datasets.

@@

The MoE is trained using datasets generated from simulations (Figure
2a). A \emph{dataset} is the minimum data required to perform 2SMR
analysis - four columns comprising the SNP-exposure and SNP-outcome
effects and standard errors, and rows corresponding to each SNP that is
used as an instrument for the exposure. This \emph{dataset} can be fed
into any of the 28 experts to obtain MR causal effects. The simulations
used to generate these datasets seek to cover a range of pleiotropic
scenarios, including where some proportion of SNPs exhibit directional
or balanced horizontal pleiotropy, or where SNPs influence confounding
variables.

We simulate two individual level datasets for which there are \(N_x\)
and \(N_y\) samples, and \(M\) SNPs, where each SNP has effect allele
frequency of \(p_m \sim U(0.05, 0.95)\). These datasets are used to
obtain the SNP effects for the exposure trait \(x\) and the outcome
trait \(y\), respectively, using the following sampling criteria:

@@

For each simulation we used linear regression to estimate the genetic
effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
\(y\) in sample 2. We then perform MR analysis in both directions,
mimicking GWAS by retaining SNPs that have \(p < 5e-8\) in sample 1 to
perform MR of \(x\) on \(y\) (the true causal direction for non-null
simulations), and retaining SNPs that have \(p < 5e-8\) in sample 2 to
perform MR of \(y\) on \(x\) (the reverse causal direction for non-null
simulations). We treat the summary data (effect sizes and standard
errors) used for estimating \(x \rightarrow y\) the summary data used
for estimating \(y \rightarrow x\) as two separate datasets Hence, for
each simulation two datasets are generated which are analysed to produce
28 MR estimates each. We performed 100,000 simulations using these
parameters, resulting in 200,000 datasets.

@@

We aim to maximise statistical power for datasets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
where \(\beta_{x,y} = 0\). To train random forest decision trees to
predict performance for a particular method
\(h(O_{w,d}, \textbf{z}_{d})\) is generated where the training set of
input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
(optimisation function) is

@@

where \(p_{m,d}\) is the p-value for method \(m\) on dataset \(d\).

@@

For each training dataset we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated dataset. For each of our 28 methods, we need to create a model
that predicts the performance of the method based on metrics generated
from a dataset. To do this, for each method we train a random decision
forest to predict that method's performance using the dataset's metrics.
The random forest approach is well suited to this problem because there
are likely to be non-continuous combinations of different metrics that
improve on prediction over, for example, a simple linear model that does
not learn about interactions.

As a simple hypothetical example - if a dataset has a single outlier but
is otherwise exhibiting no heterogeneity then the following methods
could arguably perform well:

@@

Having generated random forest decision trees for each of the 28 methods
using 133,000 of the simulations, we then applied them to the remaining
67,000 datasets to predict which method would have the highest
performance for each of the remaining datasets. Finally we compare the
performance of the method selected by the MoE against all remaining
methods. The default settings for the \texttt{randomForest} package in R
(31) were used to train the models. MR-MoE 1.0 is implemented in the
TwoSampleMR R package available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(11).

The set of MR estimates obtained from this analysis are recorded in a
Neo4j graph database. Because each association has up to 28 different
estimates, for simplicity we distill this down to a single `best
estimate' for each association using the following rules:

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered that doesn't act primarily on
the trait of interest increases (Figure 1a). We evaluated the efficacy
of the Steiger filtering approach for improving instrument selection
using 100,000 simulated datasets comprising both null and non-null
causal models. For each dataset, instruments were selected based on the
tophits strategy and the Steiger filtering strategy, and MR was
performed using 14 different methods based on instruments selected from
each of these strategies.

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 28 MR methods
and instrument selection strategies to apply to a particular dataset
based on its characteristics (Figure 2).

The ability to predict the performance for each of these methods is
shown in Supplementary table 2. The prediction \(R^2\) of whether a
dataset's status was truly null (\(\beta_{xy}=0\)) or non-null
(\(\beta_{xy} \neq 0\)) against the method's prediction of the dataset's
status, ranged between 0.04 and 0.24. The method performance prediction
was most effective for the Egger random effects model with Steiger
filtering. The dataset characteristics with the most importance for each
of the predictors differed substantially between each dataset, as well
as the frequency for which each of the methods was selected in the
testing datasets. The FDR of each method when chosen, compared to their
averages across all datasets, reduced; and likewise the true positive
rates of each method when chosen increased compared to their
simulation-wide averages.

We compared the MoE performance in the simulations against each of the
28 methods, testing to see if it outperformed all other single methods.
Figures 3a and 3b show that the MoE approach had the best general
performance, obtaining an AUC of 0.84 in terms of classifying the
simulations as being null or non-null. Notably, the next best methods
were median and mode based estimators using Steiger filtering. Under the
assumption of widespread pleiotropy it is clear that all methods suffer
from high false discovery rates.

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (32,33) and
1683 plasma protein levels (34). For the protein levels only the
instruments were available, so they could only be evaluated as exposure
phenotypes. The complex traits and diseases and metabolite levels had
complete summary data available in MR-Base, thus they be evaluated as
both exposures (if they had significant instruments) and outcomes.
Together, we evaluated 715681 relationships. The majority of these
associations could only be evaluated using fewer than 5 SNPs, and so the
Wald ratio or IVW fixed effects methods were used, but for 61029
associations the MR-MoE approach was applied.

As an example of the database we performed a look up of associations
where LDL cholesterol as measured in the GLGC consortium (35) had a
false discovery rate of 0.05 using the MoE approach. This returned 287
putative associations, amongst which 111 involved LDL cholesterol
influencing other traits and 176 involved other traits influencing LDL
cholesterol. A large proportion of these traits were metabolites, which
are dominated by lipid fractions. Filtering to exclude the metabolomic
studies ((32,33)) returned 27 associations with 23 traits (Figure 4a).
There was a strong association with coronary heart disease (0.41 log(OR)
per SD, 0.31-0.52, \(p=2.7\times10^{-12}\)). Here the IVW random effects
method was chosen after Steiger filtering, indicating a prediction of HP
being present amongst the instruments.

We also performed a look up of associations that causally influence
years of schooling (36), or that years of schooling itself causally
influences. Using a false-discovery rate of 0.05, 45 traits were
returned as having some direction of causality with years of schooling
(Figure 4b). Several of these putative relationships appear to be
plausible, for example the influence of years of schooling on lower risk
of Alzheimer's disease has recently been examined (37). However it is
clear that there are serious limitations in relying on MR as a panacea
for causal inference. For example, bi-directional causal relationships
with college completion exist because of the similarity in trait
definitions; and there are several traits (e.g.~childhood intelligence
and birth weight) which appear to be causally downstream of years of
schooling despite this being temporally impossible. Though explanations
could be conjured for this association, for example parents' schooling
influences childhood intelligence, or the genetic instruments are shared
across the two traits, an intuitive interpretation of causality cannot
be applied.

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have shown that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
are more likely to be identified than valid ones. Strategising how MR is
to be used in practice, therefore, must consider horizontal pleiotropy
to be the rule rather than the exception. Following a recent flurry in
development of MR methodology, we have devised a machine learning
approach, MR-MoE, that seeks to predict the performance of each MR
method, selecting the one which is most likely to maximise power whilst
minimising FDR for a specific dataset. The mixture of experts approach
that we present here is trained using random forest decision trees
applied to extensive and diverse simulations, and we demonstrate that it
makes substantial performance improvements over any other single method
in the presence or absence of extensive horizontal pleiotropy. As such,
our method contributes to the field of meta-learning but also to the
rapidly growing field of automatic model selection and hyper-parameter
optimization in machine learning (38).

In addition to correctly handling horizontal pleiotropy, there are many
other limitations that prevent MR from being a panacea for causal
inference. Many of the associations are biologically impossible, for
example where early life phenotypes appear to be influenced by later
stage phenotypes. Though these associations can be informative, their
interpretations as causal relationships are far from clear. Similarly,
often disease traits appear to causally relate to other phenotypes, but
GWAS is typically performed on the liability scale, hence the causal
estimate reflects not the presence or absence of disease, but the
underlying risk of disease. Again, interpretation of such associations
can be problematic.

A large proportion of the associations that were estimated used only a
single instrumental variable. Methods are emerging to attempt to
delineate between models of reverse cause, pleiotropy and multiple
causal variants in the same region (39,40). Separating vertical from
horizontal pleiotropy with a single instrument is not yet possible in
the two-sample MR framework though the use of mediation-based approaches
may be informative in the one-sample setting (7,41). Other problems can
also manifest, for example frailty effects could induce associations for
late-onset traits (42); genetic variants could strongly relate to
several phenotypes making it difficult to ascertain which of them is the
causal exposure (43); and the measured feature for which genetic
associations are known could relate in complicated ways to the
biological entity that are truly causal (2).

We also encounter a potential new problem in using machine learning
approaches to infer the correct model for a particular dataset, namely,
that the data is choosing the model. This is liable to contribute to
elevated type 1 error rates, even though we have attempted to separate
the information used for optimisation from the information used to
predict performance. The extent to which a machine based method selector
suffers from this more than a human manually inferring the result most
likely to be true is not clear. Though MR-MoE does exhibit higher type 1
error rates than are desirable, they still remain amongst the lowest
compared to all other MR methods that do not suffer from this issue.

>>>>

Gibran Hemani\(^1\), Jack Bowden\(^1\), Philip Haycock\(^1\), Jie
Zheng\(^1\), Oliver Davis\(^1\), Peter Flach\(^2\), Tom Gaunt\(^{1,*}\),
George Davey Smith\(^{1,*}\)

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  MRC Integrative Epidemiology Unit (IEU) at the University of Bristol,
  School of Social and Community Medicine, Bristol, UK
\item
  Intelligent Systems Laboratory, Department of Computer Science,
  University of Bristol
\end{enumerate}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this. Here we show that the
mechanisms that underlie horizontal pleiotropy are numerous, and that
instrument selection will be increasingly liable to selecting invalid
instruments as GWAS sample sizes continue to grow. We have developed a
mixture of experts machine learning approach (MR-MoE 1.0) that improves
on both power and false discovery rates over unselective use of any
existing methods. Using the approach, we systematically estimated the
causal effects amongst 2407 phenotypes, generating a working draft of
the causal map of the human phenome comprising the evaluation of
causality between over a million pairs of traits. Almost 90\% of causal
estimates indicated some level of horizontal pleiotropy. The causal
estimates are organised into a publicly available graph database
(\url{http://eve.mrbase.org}).

@@

The use of summary data in two-sample MR is described in detail
elsewhere (11). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of MR (that the instrument associates with the exposure) is
generally satisfied. We collect their effect sizes, standard errors and
effect alleles for the association with the exposure trait. These can be
obtained manually from complete summary data or from curated lists of
significant GWAS associations. Next, the effects of those SNPs on the
outcome need to be obtained, typically necessitating complete summary
data because these SNPs will generally have not have reached genome-wide
significance for other traits (and therefore be present in curated
catalogues).

We use the term \emph{summary-set} to refer to the minimum data required
to perform 2SMR analysis - four columns comprising the SNP-exposure and
SNP-outcome effects and standard errors, and rows corresponding to each
SNP that is used as an instrument for the exposure.

At its most simple implementation, the regression of the SNP-exposure
effect sizes against the SNP-outcome effect sizes, with greater weight
afforded to those SNPs with smaller SNP-outcome standard errors,
provides the estimate of the causal effect of the exposure on the
outcome. This is known as the inverse-variance weighted (IVW) method.

@@

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(8,12,25), based on the notion that outlying instruments are likely to
be exhibiting horizontal pleiotropy. A potential drawback of this
approach is that the outlier SNPs in a summary-set might be the only
reliable ones. This could arise, for example, for the \emph{CRP} variant
instrumenting C-reactive protein levels (19), or the \emph{SLC2A9}
variant influencing serum urate levels (26).

To avoid this, we primarily developed Steiger filtering to identify
those instruments that are likely to be arising due to reverse cause,
but it also has the potential to detect SNPs that are exhibiting
horizontal pleiotropy or primarily influencing confounders. Further
details are provided in the Online Methods.

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). In the context of this
analysis, each method is considered to be an `expert', taking a set of
SNP-exposure and SNP-outcome effect sizes and their standard errors as
inputs. Our objective is to select the expert most likely to be correct
for a specific MR analysis. An overview of the approach is shown in
Figure 2.

The mixture of experts method is a machine learning approach (27) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that
summary-set. The metrics are a mixture of regression diagnostics and are
described in Supplementary table 3.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated summary-sets. We trained the gating function using random
forest learning algorithms that seek to identify sectors of the
parameter space that are most likely to return accurate estimates for a
particular expert. Figure 2a illustrates how the gating function is
trained using simulated data. New summary-sets can then be applied to
the trained mixture of experts (MoE) model (Figure 2b). We call this
implementation MR-MoE 1.0. Full details are provided in the Online
Methods. MR-MoE 1.0 is implemented in the TwoSampleMR R package
available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(11).

The set of MR estimates obtained from this analysis, and the
summary-sets used to generate them, are deposited in a Neo4j graph
database. Here nodes representations exist for traits and SNPs; and
relationship representations exist for SNP-trait associations and
trait-trait MR estimates. Because each trait-trait association has up to
28 different MR estimates, for simplicity we also distill this down to a
third relationship type comprising a single `best estimate' for each
trait-trait association using the following rules:

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered that doesn't act primarily on
the trait of interest increases (Figure 1a). We evaluated the efficacy
of the Steiger filtering approach for improving instrument selection
using 100,000 simulated summary-sets comprising both null and non-null
causal models. For each summary-set, instruments were selected based on
the tophits strategy and the Steiger filtering strategy, and MR was
performed using 14 different methods based on instruments selected from
each of these strategies.

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 28 MR methods
and instrument selection strategies to apply to a particular summary-set
based on its characteristics (Figure 2).

The ability to predict the performance for each of these methods is
shown in Supplementary table 2. The prediction \(R^2\) of whether a
summary-set's status was truly null (\(\beta_{xy}=0\)) or non-null
(\(\beta_{xy} \neq 0\)) against the method's prediction of the
summary-set's status, ranged between 0.04 and 0.24. The method
performance prediction was most effective for the Egger random effects
model with Steiger filtering. The summary-set characteristics with the
most importance for each of the predictors differed substantially
between each summary-set, as well as the frequency for which each of the
methods was selected in the testing summary-sets. The FDR of each method
when chosen, compared to their averages across all summary-sets,
reduced; and likewise the true positive rates of each method when chosen
increased compared to their simulation-wide averages.

We compared the MoE performance in the simulations against each of the
28 methods, testing to see if it outperformed all other single methods.
Figures 3a and 3b show that the MoE approach had the best general
performance. Estimating the area under the receiver operator curve gave
0.84 for the MoE approach in terms of classifying the simulations as
being null or non-null. Notably, the next best methods were median and
mode based estimators using Steiger filtering. Under the assumption of
widespread and diverse pleiotropic effects it is clear that all methods
suffer from high false discovery rates (Figure 3a).

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (28,29) and
1683 plasma protein levels (30). For the protein levels only the
instruments were available, so they could only be evaluated as exposure
phenotypes. The complex traits and diseases and metabolite levels had
complete summary data available in MR-Base, thus they be evaluated as
both exposures (if they had significant instruments) and outcomes.
Together, we evaluated 715681 relationships. The majority of these
associations could only be evaluated using fewer than 5 SNPs, and so the
Wald ratio or IVW fixed effects methods were used, but for 61029
associations the MR-MoE approach was applied.

As an example of the database we performed a look up of associations
where LDL cholesterol as measured in the GLGC consortium (31) had a
false discovery rate of 0.05 using the MoE approach. This returned 287
putative associations, amongst which 111 involved LDL cholesterol
influencing other traits and 176 involved other traits influencing LDL
cholesterol. A large proportion of these traits were metabolites, which
are dominated by lipid fractions. Filtering to exclude the metabolomic
studies ((28,29)) returned 27 associations with 23 traits (Figure 4a).
There was a strong association with coronary heart disease (0.41 log(OR)
per SD, 0.31-0.52, \(p=2.7\times10^{-12}\)). Here the IVW random effects
method was chosen after Steiger filtering, indicating a prediction of HP
being present amongst the instruments.

We also performed a look up of associations that causally influence
years of schooling (32), or that years of schooling itself causally
influences. Using a false-discovery rate of 0.05, 45 traits were
returned as having some direction of causality with years of schooling
(Figure 4b). Several of these putative relationships appear to be
plausible, for example the influence of years of schooling on lower risk
of Alzheimer's disease has recently been examined (33). However it is
clear that there are serious limitations in relying on MR as a panacea
for causal inference. For example, bi-directional causal relationships
with college completion exist because of the similarity in trait
definitions; and there are several traits (e.g.~childhood intelligence
and birth weight) which appear to be causally downstream of years of
schooling despite this being temporally impossible. Though explanations
could be conjured for this association, for example parents' schooling
influences childhood intelligence, or the genetic instruments are shared
across the two traits, an intuitive interpretation of causality cannot
be applied.

The trend of increasing GWAS sample sizes continues, but while the
opportunity that this affords MR to be furnished with more instruments
is typically welcomed, here we have shown that it is invalid
instruments, exhibiting horizontal pleiotropy or reverse causation, that
are more likely to be identified than valid ones. Strategising how MR is
to be used in practice, therefore, must consider horizontal pleiotropy
to be the rule rather than the exception. Following a recent flurry in
development of MR methodology, we have devised a machine learning
approach, MR-MoE, that seeks to predict the performance of each MR
method, selecting the one which is most likely to maximise power whilst
minimising FDR for a specific summary-set. The mixture of experts
approach that we present here is trained using random forest decision
trees applied to extensive and diverse simulations, and we demonstrate
that it makes substantial performance improvements over any other single
method in the presence or absence of extensive horizontal pleiotropy. As
such, our method contributes to the field of meta-learning but also to
the rapidly growing field of automatic model selection and
hyper-parameter optimization in machine learning (34).

One advantage to generating a method that consolidates across many
methods is that it can prevent cherry picking the most appealing method
based on their results. But in addition to correctly handling horizontal
pleiotropy, there are many other limitations that prevent MR from being
a panacea for causal inference. Many of the associations are
biologically impossible, for example where early life phenotypes appear
to be influenced by later stage phenotypes. Though these associations
can be informative, their interpretations as causal relationships are
far from clear. Similarly, often disease traits appear to causally
relate to other phenotypes, but GWAS is typically performed on the
liability scale, hence the causal estimate reflects not the presence or
absence of disease, but the underlying risk of disease. Again,
interpretation of such associations can be problematic.

A large proportion of the associations that were estimated used only a
single instrumental variable. Methods are emerging to attempt to
delineate between models of reverse cause, pleiotropy and multiple
causal variants in the same region (35,36). Separating vertical from
horizontal pleiotropy with a single instrument is not yet possible in
the two-sample MR framework though the use of mediation-based approaches
may be informative in the one-sample setting (7,37). Other problems can
also manifest, for example frailty effects could induce associations for
late-onset traits (38); genetic variants could strongly relate to
several phenotypes making it difficult to ascertain which of them is the
causal exposure (39); and the measured feature for which genetic
associations are known could relate in complicated ways to the
biological entity that are truly causal (2).

We also encounter a potential new problem in using machine learning
approaches to infer the correct model for a particular summary-set,
namely, that the data is choosing the model. This is liable to
contribute to elevated type 1 error rates, even though we have attempted
to separate the information used for optimisation from the information
used to predict performance. The extent to which a machine based method
selector suffers from this more than a human manually inferring the
result most likely to be true is not clear. Though MR-MoE does exhibit
higher type 1 error rates than are desirable, they still remain amongst
the lowest compared to all other MR methods that do not suffer from this
issue.

The field of MR is evolving. The advent of GWAS databases (9,40) and
automated 2SMR (11) has trivialised the analytical aspect of
investigating specific causal hypotheses. Despite the limitations
described above, the construction of a causal graph of `everything
versus everything' does have appeal. First, though causality is not
guaranteed by MR, it can still be highly informative for confirming or
negating specific hypotheses. Second, it paves the way for new
approaches to search for novel putative associations and improve
reliability in single analyses. The potential to exploit GWAS summary
data within the properties of graph databases to aid with both of these
endeavours could be transformative in biological research.

\subsection{Acknowledgements}\label{acknowledgements}

This work was supported by the UK Medical Research Council
(MC\_UU\_12013/1).

\subsection{Online methods}\label{online-methods}

\subsubsection{Steiger filtering}\label{steiger-filtering-1}

An approach to inferring the causal direction between phenotypes
recently developed (7) uses the following basic premise. If trait \(A\)
causes trait \(B\) then \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\) because
the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This will be
true under most circumstances, but some parameters of measurement error
or unmeasured confounding can lead to the inequality reversing
direction, as has been explored previously (7).

The Steiger test is applied to each variant in turn and we exclude any
\(g_{A}\) for which \(cor(g_{i}, A)^2 > cor(g_{i}, B)^2\), indicating
that it is unlikely to primarily associate with \(B\) relative to \(A\).
Similarly, for SNPs that influence confounders of \(A\) and \(B\) or
exhibit horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant. Hence we also exclude any \(g_{A}\) for which the
\(cor(g, x)\) and \(cor(g, y)\) are not significantly different at an
arbitrary threshold of \(p > 0.05\).

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2f(1-f)\), where \(\beta\) is the log odds ratio and
\(f\) is the allele frequency of the SNP in the population (41). \(f\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\(f_{cc}\), and number of cases \(N_1\) and controls \(N_0\). A Z-score
to test the hypothesis that \(cor(g, x)\) and \(cor(g, y)\) are
significantly different can be generated using the Steiger test for
correlated correlations.

\subsubsection{Mixture of experts implementation for
MR}\label{mixture-of-experts-implementation-for-mr}

The mixture of experts approach seeks to use simulations to learn the
circumstances in which each of the 28 MR methods is likely to perform
most accurately. Having trained the model on simulations, new
summary-sets can be applied to obtain causal estimates from the method
that is predicted to be the most reliable for that summary-set.

It should be noted that in the original hierarchical mixture of experts
approach of (27) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (42--44) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the summary-sets.

@@

The MoE is trained using summary-sets generated from simulations (Figure
2a). Each \emph{summary-set} can be fed into any of the 28 experts to
obtain MR causal effects. The simulations used to generate these
summary-sets seek to cover a range of pleiotropic scenarios, including
where some proportion of SNPs exhibit directional or balanced horizontal
pleiotropy, or where SNPs influence confounding variables.

We simulate two individual level datasets for which there are \(N_x\)
and \(N_y\) samples, and \(M\) SNPs, where each SNP has effect allele
frequency of \(f_m \sim U(0.05, 0.95)\). These datasets are used to
obtain the SNP effects for the exposure trait \(x\) and the outcome
trait \(y\), respectively, to create summary-sets, using the following
sampling criteria:

@@

For each simulation we used linear regression to estimate the genetic
effect of each SNP \(M\) on \(x\) in sample 1, and each SNP \(M\) on
\(y\) in sample 2. We then perform MR analysis in both directions,
mimicking GWAS by retaining SNPs that have \(p < 5e-8\) in sample 1 to
perform MR of \(x\) on \(y\) (the true causal direction for non-null
simulations), and retaining SNPs that have \(p < 5e-8\) in sample 2 to
perform MR of \(y\) on \(x\) (the reverse causal direction for non-null
simulations). We treat the summary data (effect sizes and standard
errors) used for estimating \(x \rightarrow y\) the summary data used
for estimating \(y \rightarrow x\) as two separate summary-sets. Hence,
for each simulation two summary-sets are generated which are analysed to
produce 28 MR estimates each. We performed 100,000 simulations using
these parameters, resulting in 200,000 summary-sets.

@@

We aim to maximise statistical power for summary-sets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for
summary-sets where \(\beta_{x,y} = 0\). To train random forest decision
trees to predict performance for a particular method
\(h(O_{w,d}, \textbf{z}_{d})\) is generated where the training set of
input metrics for summary-set \(d\) is \(\textbf{z}_{d}\) and the
response (optimisation function) is

@@

where \(p_{m,d}\) is the p-value for method \(m\) on summary-set \(d\).

@@

For each training summary-set we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 3), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated summary-set. For each of our 28 methods, we need to create a
model that predicts the performance of the method based on metrics
generated from a summary-set. To do this, for each method we train a
random decision forest to predict that method's performance using the
summary-set's metrics. The random forest approach is well suited to this
problem because there are likely to be non-continuous combinations of
different metrics that improve on prediction over, for example, a simple
linear model that does not learn about interactions.

As a simple hypothetical example - if a summary-set exhibits a single
outlier but is otherwise exhibiting no heterogeneity then the following
methods could arguably perform well:

@@

Having generated random forest decision trees for each of the 28 methods
using 133,000 of the simulations, we then applied them to the remaining
67,000 summary-sets to predict which method would have the highest
performance for each of the remaining summary-sets. Finally we compare
the performance of the method selected by the MoE against all remaining
methods. The default settings for the \texttt{randomForest} package in R
(45) were used to train the models. MR-MoE 1.0 is implemented in the
TwoSampleMR R package available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}
(11).
